Phylogenetics and the human microbiome.

### 2013-10-02 sbreview.tex
A few small fixes from Bastien.

In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perpsective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

@@

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except for in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, although multiplex sequencing is known to have biases of its own.

@@

Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper.

@@

The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as it such errors are especially problematic there.
With the exception of \citet{sharpton2011phylotu}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Nevertheless, the work that has been done to infer horizontal gene transfer in the human microbiome has revealed interesting results.
\citet{hehemann2010transfer} found that a gene found in seaweed has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.

@@

Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
There are two primary such projects: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nt,
% SILVA
% 1543 / 42284
% .96353230536373096207
the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
The GreenGenes tree is updated by running FastTree from scratch for every release.
There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.

In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a very nice interactive tree-based genome browser.
On a much smaller scale, there curated reference sets that are microbiome body-site specific \citep{chen2010human,griffen2011core,srinivasan2012bacterial}

@@

One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
These two extremes are each each useful in different parameter regimes, but something between the two would be helpful as well.
For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.

@@

In a similar vein, assessing the significance level an observed difference between communities is not straightforward and poses difficult problems.
The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
Such non-independence can lead to incorrect rejection of the null hypothesis.
Imagine, for example, that we have a random observation process on the tree equipped with some collection of ``base observations'' at the leaves.
Each sample from the process takes a random subset of those base observations and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
If the set of base observations is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
This would be a false positive.
Even basic definitions pose a challenge here: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.

>>>>

In this paper we will be primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

@@

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that a collection of sequences need not be representative of the communities from which they were derived \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts \citep{felsenstein1985phylogenies} method to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except in the inference of operational taxonomic units as described below.

\subsection{Metagenomes}
As described above ``metagenome" means that data is sheared randomly across the genome rather than amplified from a specific location, and thus the genetic region of a read is unknown in addition to what organism it was derived from.
Because metagenomes do not proceed through an amplification step, it does not have the same PCR primer biases as a marker gene survey, however extraction efficiency concerns remain and multiplex sequencing is known to have biases of its own.

@@

Metagenomic data is often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper, as is ``metatranscriptomics": the sequencing of mRNA in bulk.

@@

The centrality of the OTU concept can be seen by the fact that the contingency table of OTU observations is considered to be the fundamental data type for 16S studies \citep{mcdonald2012biological}, or that methods have been devised to find OTUs from non-overlapping sequences \citep{sharpton2011phylotu}.
A significant amount of effort has been made to distinguish sequencing error in environmental samples from true rare variants; much of this work has played out in the OTU inference literature \citep{quince2009accurate,quince2011removing,bragg2012fast,edgar2013uparse} as such errors are especially problematic there.
With the exception of \citet{sharpton2011phylotu}, OTU inference is not considered to be a phylogenetic problem but rather something to be performed before phylogenetic inference begins.

@@

Nevertheless, the work that has been done to infer horizontal gene transfer in the human microbiome has revealed interesting results.
\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.

@@

Considerable effort goes to the creation of large curated alignments and phylogenetic trees on 16S.
There are two major projects to do so: one is the SILVA database \citep{pruesse2007silva,quast2013silva}, and the other is the GreenGenes database \citep{desantis2006greengenes,mcdonald2011improved}.
Because of the high rate of insertion and deletion of nucleotides in 16S, these alignments have a high percentage of gap.
Taking the length of 16S to be 1543 nt,
% SILVA
% 1543 / 42284
% .96353230536373096207
the 479,726 sequence SILVA reference alignment version 115 is over 96\% gap,
% GreenGenes
% 1543 / 7682
% .79914084873730799272
while the 1,262,986 sequence GreenGenes 13\_5 alignment has is almost 80\% gap.
The SILVA-associated 'all-species living tree' project \citep{yarza2008all} started with a tree inferred by maximum likelihood and has been continually updated  by inserting sequences via parsimony.
The GreenGenes tree is updated by running FastTree from scratch for every release.
There appears to be a commonly held belief that FastTree in particular works well even with such gappy alignments \citep[e.g.][]{sharpton2011phylotu}.

In addition to these 16S-based resources, the MicrobesOnline resource \citep{dehal2010microbesonline} offers a very nice interactive tree-based genome browser.
On a much smaller scale, there are microbiome body-site specific reference sets \citep{chen2010human,griffen2011core,srinivasan2012bacterial}

@@

One clear challenge is to fill the gap between on one hand complete \emph{de novo} tree inference versus sequence insertion or placement that leaves the ``reference tree'' fixed.
These two extremes are each useful in different parameter regimes, but something between the two would be helpful as well.
For example, sequence data is continually being added to large databases, motivating methods that could continually update trees with this new sequence data while allowing the previous tree to change according to this new information.
% Surprising that methods such as DCM \citep{huson1999disk} have not gained traction.

@@

In a similar vein, assessing the significance level of an observed difference between communities is not straightforward and poses difficult problems.
The randomization of group membership commonly used in combination with UniFrac to determine significance does not have appropriate properties when in the regime of incomplete sampling with non-independent observations, which is certainly the correct regime for surveys and metagenomes.
Such non-independence can lead to incorrect rejection of the null hypothesis.
Imagine, for example, that we have a random observation process on the tree equipped with some collection of ``base observations'' at the leaves.
Each sample from the process takes a random subset of those base observations and then throws down some number of reads for each observation in that set, the number of which has mean significantly greater than 1.
If the set of base observations is large compared to the number of sample observations, then two draws will always appear significantly different even though they are from the same underlying process.
This would be a false positive.
Even basic definitions pose a challenge here: the question of whether two communities are the ``same'' and ``different'' probably needs to be approached from the perspective of ecosystem modeling.
